---
author: "Julia McCartney"
date: "Due 9/18/2020"
output: html_document
---
# Homework 1: Intro to R {.tabset}

#### Julia McCartney  
#### Due 9/18/2020

Here is the link to this file on [GitHub](https://github.com/julia-mccartney/McCartney_Biol_607/blob/master/homework/01_McCartney_Julia_2020.Rmd)  


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

Load quakes with data(quakes). Show what’s there with str() and summary().


```{r }
data("quakes")

str(quakes)
```

```{r, eval = FALSE}
summary(quakes)
```


```{r, echo=FALSE}

knitr::kable(summary(quakes))
```


## Question 2

Show the entirety of the column long.

```{r}
quakes$long
```


## Question 3

Hey, another useful function - unique()! Apply it to a vector, and you can see what are all of the unique values. It’s great for really digging into a problematic vector.

What unique stations are there? Use length() with unique() to determine how many stations there are.

```{r}
length(unique(quakes$stations))
```

There are 102 unique stations.

## Question 4

Using range(), what is the range of depths where quakes occur?

```{r}
range(quakes$depth)
```

The earthquakes in this dataset occurred at depths between 40 and 680 km.

## Extra Credit

E.C. Where was the earthquake of largest magnitude found? You’ll need to use come comparisons and the max() function for this in your indices instead of a number!

```{r, eval = FALSE}
max_mag <- max(quakes$mag)

quakes[quakes$mag %in% max_mag,]
```


```{r, echo=FALSE}
max_mag <- max(quakes$mag)

knitr::kable(quakes[quakes$mag %in% max_mag,])

```


The largest quake in this dataset was at latitude -15.56, longitude 167.62.   
  


```{r, echo=FALSE}
library(RgoogleMaps)


BGB =list(lat =-15.56360, lon =167.62120)
par(pty="s")
map1 = plotmap(BGB$lat, BGB$lon, zoom = 9, maptype = "terrain", size = c(640,
640))

```

  
Map created using the package [RgoogleMaps](https://cran.r-project.org/web/packages/RgoogleMaps/RgoogleMaps.pdf)


